Supporting Information for: Phonon polaritonics in two-dimensional materials

### 2019-03-18 supp.tex
supp

\documentclass[aps,prb,twocolumn,
	           groupedaddress,superscriptaddress,
               amsfonts,amssymb,amsmath,floatfix,
	           citeautoscript]{revtex4-1}

@@

% ----- COMMENTS AND META-ANNOTATIONS -----
\usepackage{textcomp} % for \textrightarrow
\usepackage{xifthen}
\usepackage{etoolbox}
\newboolean{togglecomments}
\newboolean{togglechanges} 
\newboolean{toggleadditioncolor}

% toggle to true to see comments (otherwise hidden)
\setboolean{togglecomments}{true}  
% toggle to false to see mixed versions (otherwise edits are shown exclusively)
\setboolean{togglechanges}{false} 
% toggle to true to see comments (otherwise hidden)
\setboolean{toggleadditioncolor}{true}

\newcommand{\comment}[2]{%
    \ifbool{togglecomments}%
    {\textcolor{blue!70!black}{\small\textsf{%
    \textsuperscript{\textsc{\textsf{\MakeLowercase{#1}}}}%
    [#2]}}} % if true, show comments
    {}}     % if false, do nothing
\newcommand{\swap}[2]{\ifbool{togglechanges}
    {#2}  % TC-only version
    {\textcolor{red!70!black}{[#1]}\text{\textrightarrow{}}\textcolor{green!50!black}{[#2]}}}
\newcommand{\remove}[1]{\ifbool{togglechanges}
    {}    % TC-only version
    {\textcolor{red!70!black}{#1}}}
\newcommand{\inset}[1]{\ifbool{togglechanges}
    {{}#1}  % TC-only version
    {{}{\color{green!50!black}#1}}}
\newcommand{\citeremind}[1]{%
	\unskip%
    \textcolor{blue!75!black!80!yellow}{${}^\blacksquare$%
	\ifthenelse{\isempty{#1}}{}{\textsuperscript{\tiny\textsf{#1}}}%
	}\xspace}
\newcommand{\addition}[1]{\ifbool{toggleadditioncolor}
	{\textcolor{blue!65!black}{{}#1}}  % colored version
	{#1}}
\newcommand{\FIXME}[1]{\textcolor{red!70!black}{[#1]}}
\newcommand{\optional}[1]{\textcolor{yellow!70!gray}{#1}}

@@

% --- TITLE & AUTHORSHIP ---
\title{Phonon polaritonics in two-dimensional materials}

@@

\date{\today}

% --- ABSTRACT ---
\begin{abstract}
    Extreme confinement of electromagnetic energy by phonon polaritons holds the promise of strong and new forms of control over the dynamics of matter. 
    To bring such control to the atomic-scale limit, it is important to consider phonon polaritons in two-dimensional (2D) systems. Recent studies have pointed out that in 2D, splitting between longitudinal and transverse optical (LO and TO) phonons is absent at the $\Gamma$ point, even for polar materials.  
    Does this lack of LO--TO splitting imply the absence of a phonon polariton in polar monolayers? 
    Here, we derive a first-principles expression for the conductivity of a polar monolayer specified by the wavevector-dependent LO and TO phonon dispersions. In the long-wavelength (local) limit, we find a universal form for the conductivity in terms of the LO phonon frequency at the $\Gamma$ point,
    its lifetime, and the group velocity of the LO phonon. Our analysis reveals that the phonon polariton of 2D is simply the LO phonon of the 2D system. For the specific example of hexagonal boron nitride (hBN), we estimate the confinement and propagation losses of the LO phonons, finding that high confinement and reasonable propagation quality factors coincide in regions which may be difficult to detect with current near-field optical microscopy techniques. Finally, we study the interaction of external emitters with two-dimensional hBN nanostructures, finding extreme enhancement of spontaneous emission due to coupling with localized 2D phonon polaritons, and the possibility of multi-mode strong and ultra-strong coupling between an external emitter and hBN phonons. This may lead to the design of new hybrid states of electrons and phonons based on strong coupling.

\end{abstract}

@@

% --- INTRODUCTION ---
Phonon polaritons, hybrid quasiparticles of photons and optical phonons supported in polar materials, hold promise for nanoscale control of electromagnetic fields at mid-infrared and terahertz frequencies. Qualitatively, phonon polaritons share many features with plasmon polaritons in conductors. Recently, it has been shown that phonon polaritons enable confinement of light to volumes  $\sim10^6$ times smaller than that of a diffraction-limited photon in free-space~\cite{caldwell2013low,xu2014mid,caldwell2014sub,dai2014tunable,tomadin2015accessing,yoxall2015direct,li2015hyperbolic,dai2015subdiffractional,dai2015graphene,caldwell2015low,li2016reversible,Basov:2016,basov2017towards,low2017polaritons,giles2017ultra,li2018infrared,ma2018plane}. Due to this remarkable confinement and their relatively high lifetimes---around picoseconds---phonon polaritons open new opportunities for vibrational spectroscopy~\cite{autore2018boron}, radiative heat transfer~\cite{hillenbrand2002phonon}, and control of dynamics in quantum emitters~\cite{kumar2015tunable,rivera2017making,kurman2018control,flick2018strong}.

Thus far, extreme confinement of phonon polaritons has been achieved by the use of thin-films (or nanostructuring), which shrink the in- and out-of-plane wavelength of polaritons with decreasing feature size (such as the film thickness)~\cite{dai2014tunable,dubrovkin2018ultra}. A monolayer is the ultimate limit of this effect, making it critical to have a fundamental understanding of the optical response of 2D polar materials~\cite{thygesen2017calculating}. Concerning the optical response, the transition from three-dimensional (3D) to two-dimensional (2D) polar materials is nontrivial, however, since in a polar monolayer, the LO--TO splitting that gives rise to phonon polaritons in 3D is absent at the $\Gamma$ point~\cite{sanchez2002vibrational,mele2002electric,serrano2007vibrational,sohier2017breakdown}. This raises a fundamental question about the nature of  electromagnetic modes in polar monolayers.

We now develop a framework with quantitative, concrete examples that enables a first-principles understanding of phonon polaritons in 2D materials. 
We do so by deriving a universal form for the conductivity of a polar monolayer,  which depends solely on the LO and TO phonon frequencies---and their dispersion with momentum---in the 2D system. Using parameters from Ref.~\citenum{sohier2017breakdown} for the canonical 2D polar monolayer---hexagonal boron nitride (hBN)---we present the confinement and propagation losses of the 2D phonon polariton modes, identifying the frequency region where they should be most easily detected. Finally, we find that these modes enable extreme light-matter interaction between emitters and polar materials, showing that for atom-like emitters, their spontaneous decay can be enhanced by up to eight orders of magnitude through the emitter--LO phonon coupling. For an infrared emitter with a sufficiently high free-space radiative decay rate (${\gtrsim}\, \SI{1e6}{\per\s}$), we find that the associated linewidth of the emitter is comparable to the spacing between different phonon polaritonic resonances of an hBN nanostructure. This suggests the possibility of realizing the multi-mode strong coupling and ultra-strong coupling regimes of quantum electrodynamics in a 2D hBN platform. Our results for hBN are particularly relevant due its widespread use in 2D van der Waals heterostructures. In addition to providing functionality as a layer which improves the electrical and optical properties of other 2D materials, \eg graphene, our results suggest that in these heterostructures, hBN layers could provide a mid-infrared platform for nanophotonics and quantum optics. While we focus on hBN in this manuscript, the salient features of our findings apply to other polar monolayers as well.

\section{Optical response of optical phonons in two-dimensions}
In this section, we develop a theory of electromagnetic response due to optical phonons in 2D systems. The key response function of interest is the conductivity of the monolayer. To that end, we consider the response of the ions of the monolayer due to an electric potential $\phi$. For that case, the interaction Hamiltonian is
\begin{equation}
    H_{\mathrm{int}} = \int \dd^2x ~\rho \phi = -\int \dd^2x~ (\nabla\cdot\mathbf{P})\phi,
    \label{eq:coupling}
\end{equation} 
with $\rho$ the induced charge density and $\mathbf{P}$ the induced polarization density associated with the ionic motion. Note that boldfaced quantities refer to vectors or tensors as appropriate. Within linear response theory, the polarization density can be straightforwardly evaluated from the displacement  $\mathbf{u}_{\kappa}$ of every atom $\kappa$ within the unit cell. Specifically, to first order, the polarization density is 
\begin{equation}
\mathbf{P} - \mathbf{P}^0 = \sum_\kappa(\mathbf{u}_\kappa\cdot\boldsymbol{\nabla}_{\mathbf{u}_\kappa})\mathbf{P} \equiv \frac{1}{\Omega}\sum_\kappa\mathbf{Z}_{\kappa}\mathbf{u}_{\kappa},
\end{equation} 
where $\mathbf{Z}_\kappa \equiv \Omega\boldsymbol{\nabla}_{\mathbf{u}_\kappa}\mathbf{P}$ is the Born effective charge tensor of ion $\kappa$ and $\Omega$ is the unit cell area. $\mathbf{P}^0$ is the equilibrium polarization in the absence of displacements, which is zero here. With this relation between polarization and ionic displacements, the interaction Hamiltonian in Equation~\eqref{eq:coupling} couples the scalar potential and the ionic displacements.  We consider the response of the monolayer to a potential of the form $\phi(\mathbf{r}) = \phi(\mathbf{q},\omega)\e^{\iu\mathbf{q}\cdot\mathbf{r}-\iu\omega t}$, where $\mathbf{q}$ is a 2D wavevector in the plane of the monolayer. Such a potential corresponds to a longitudinal electric field  $\mathbf{E}(\mathbf{r}) = \iu\mathbf{q}\phi(\mathbf{r})$.

In what follows, we assume the validity of the random-phase approximation (RPA) in calculating Coulombic interactions between ions in the polar lattice. Within the RPA, these Coulombic interactions are accounted for by taking the induced polarization $\mathbf{P}(\mathbf{q},\omega)$ to be proportional \emph{total} electric field, $\mathbf{E}_{\mathrm{tot}}(\mathbf{q},\omega)$, defined to be the sum of the externally applied electric field and the electric field created by the induced polarization. The polarization and total field are connected by the polarization-polarization response (tensor) function  $\boldsymbol{\Pi}(\mathbf{q},\omega)$ via
\begin{equation}
\mathbf{P}(\mathbf{q},\omega)  = \epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}_{\mathrm{tot}}(\mathbf{q},\omega),
\end{equation}
The polarization-polarization response function is related to the conductivity via the relation $\boldsymbol{\sigma}(\mathbf{q},\omega) = -\iu\omega\epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)$. From the Kubo formula, it follows that the conductivity is
\begin{equation}\label{eq:2dsusceptibility}
    \boldsymbol{\sigma}(\mathbf{q},\omega) =  \frac{-\iu \omega}{ \Omega\mathcal{Z}}\sum\limits_{m,n}\frac{\mathbf{P}_{mn}(\mathbf{q})\otimes\mathbf{P}_{nm}(\mathbf{q})}{\hbar\omega + E_{nm}+\iu 0^+}\Big(\e^{-\beta E_m}-\e^{-\beta E_n} \Big),
\end{equation}
where $m,n$ are eigenstates in the phononic Fock space of the monolayer, $\mathbf{P}_{mn}(\mathbf{q}) \equiv \sum_\kappa \mathbf{Z}_\kappa\langle m|\mathbf{u}_\kappa(\mathbf{q})|n\rangle$ are matrix elements of the polarization associated with phonon modes (where $\mathbf{u}_\kappa(\mathbf{q})$ is the Fourier transform of the phonon displacement operator), $E_{m}$ ($E_n$) is the energy of state $m$ ($n$), $\beta \equiv 1/k_{\text{\textsc{B}}}T$ is the inverse temperature, and $\mathcal{Z}$ is the grand partition function. We now evaluate the contribution of optical phonons to the polarization-polarization response in the low-temperature limit $T \ll \hbar\omega_{\mathrm{ph}}/k_{\mathrm{B}}$ with $\omega_{\mathrm{ph}}$ a characteristic optical phonon frequency.  Considering the long-wavelength (small wavevector) limit, and taking a material with long-wavelength isotropy, such as hBN, we only have to consider the $qq$-component in the response tensor, where $qq$ denotes a pair of directions parallel to the wavevector. Denoting $\sigma_{qq}$ as simply $\sigma$, we find that the conductivity is given by~\cite{rivera2018ab}: 
\begin{equation}
    \sigma(\mathbf{q},\omega) = \frac{-\iu\omega}{\hbar\Omega} \frac{2\omega_{\mathbf{q},\mathrm{L}}}{\omega^2_{\mathbf{q},\mathrm{L}}-\omega^2-\iu\omega\tau^{-1}}|\hat{\mathbf{q}}\cdot\langle 1_{\mathbf{q},\mathrm{L}}|\mathbf{P}(\mathbf{q})|0_{\mathbf{q},L}\rangle|^2,
    \label{eq:polpolresponse}
\end{equation}
where L-subscripts denote longitudinal polarization, $|0_{\mathbf{q},\mathrm{L}}\rangle$ ($|1_{\mathbf{q},\mathrm{L}}\rangle$) denotes a state with no (one) longitudinal phonon of wavevector $\mathbf{q}$, and $\hat{\mathbf{q}}$ denotes a unit vector in the direction of $\mathbf{q}$. We have also phenomenologically included the phonon dissipation rate $\tau^{-1}$, consistently with  a relaxation-time prescription.  The frequency $\omega_{\mathbf{q},\mathrm{L}}$ in the denominator, as in the case of bulk phonons, is the frequency of the longitudinal phonon of wavevector $\mathbf{q}$ prior to considering LO--TO splitting~\cite{BornHuang:1954} (and near the $\Gamma$ point is approximately equal to the TO phonon frequency). This is consistent with the fact that LO--TO splitting is a collective effect arising from Coulomb interactions and the fact that the equation above represents a single-particle susceptibility. Coulomb interactions are accounted for in the random phase approximation, and to include them in the single-particle response amounts to an uncontrolled double-counting.

Next, we express the polarization matrix element in Equation~\eqref{eq:polpolresponse} in terms of the Born effective charges of the monolayer and the phonon displacement eigenvectors. Considering the longitudinal phonon contribution to the second-quantized ionic displacement, as in Ref.~\citenum{srivastava1990physics}, we find that the conductivity within the RPA is given by
\begin{equation}
    \sigma(\mathbf{q},\omega) = -\frac{\iu\omega }{\Omega}\frac{\Big|\hat{\mathbf{q}}\cdot\sum\limits_{\kappa}\mathbf{Z}_{\kappa}\boldsymbol{\eta}_{\kappa} \Big|^2}{\omega^2_{\mathbf{q},\mathrm{L}}-\omega^2-\iu\omega\tau^{-1}}.
    \label{eq:conductivity_simplified}
\end{equation} 
We have defined scaled eigendisplacements $\boldsymbol{\eta}_{\kappa\mathbf{q}}\equiv \frac{\hat{\mathbf{e}}_{\kappa\mathbf{q},\mathrm{L}}}{\sqrt{M_{\kappa}}}$, where $\hat{\mathbf{e}}_{\kappa\mathbf{q},\mathrm{L}}$ is the unit-normalized polarization vector of atom $\kappa$ in the unit cell oscillating according to a longitudinal phonon of wavevector $\mathbf{q}$ and $M_{\kappa}$ is the mass of atom $\kappa$.

While the conductivity is the main electromagnetic quantity of interest for electrodynamics applications, we briefly state the form of the (2D) permittivity, as its zeros immediately yield the longitudinal modes of the system, which are the LO phonons. The permittivity within the RPA, denoted $\epsilon_{\textrm{RPA}}$ is related to the polarization-polarization response function via~\cite{jablan2009plasmonics} $\epsilon_{\textrm{RPA}} = \epsilon_{\mathrm{env}} + \frac{1}{2}q\Pi(q,\omega)$ and the conductivity via $\epsilon_{\textrm{RPA}} = \epsilon_{\mathrm{env}} + \frac{\iu q\sigma(q,\omega)}{2\omega\epsilon_0}$. Here, $\epsilon_{\mathrm{env}}$ is the average permittivity of the bulk above and below the monolayer, and is added to take into account the polarization arising from these bulk materials. Note that we have neglected any intrinsic high-frequency screening in the monolayer itself, which is only relevant for wavevectors comparable to the inverse layer spacing between monolayers. When considering non-local corrections to the conductivity at these large wavevectors, these must be taken into account~\cite{sohier2016two,thygesen2017calculating,sohier2017breakdown}. Based on Equation~\eqref{eq:conductivity_simplified}, the zeros $\omega_{\mathbf{q}}$ of the RPA dielectric function satisfy:
\begin{equation}
    \omega^2_{\mathbf{q}} - \omega^2_{\mathrm{TO}} = \frac{V(q)}{e^2}\frac{1}{\Omega}q^2\Big|\hat{\mathbf{q}}\cdot\sum\limits_{\kappa}\mathbf{Z}_{\kappa}\boldsymbol{\eta}_{\kappa}\Big|^2,
    \label{eq:zerosofresponse}
\end{equation}
  where $V(q)$ is the Coulomb interaction in Fourier space, which in two dimensions, is given by $V(q) = \frac{e^2}{2\epsilon_0\epsilon_{\mathrm{env}}q}$. Given that the zeros of the dielectric function are associated with longitudinal modes, one expects that $\omega_{\mathbf{q}}$ is in fact the frequency of the LO phonon mode. This is consistent with the result of Ref.~\citenum{sohier2017breakdown}, in which \relax{{}\ }\citeauthor{sohier2017breakdown} show that in 2D polar materials, the extra restoring forces on LO phonons relative to TO phonons, due to the Coulomb interaction, lead to a wavevector-dependent LO--TO splitting and zero LO--TO splitting at the $\Gamma$ point of the Brillouin zone.

Given these results, we now re-express the conductivity explicitly in terms of the 2D phonon dispersion, and derive a universal form for the conductivity in the local ($q\rightarrow 0$) limit specified in terms of three parameters: the LO phonon frequency at the $\Gamma$ point (i.e., $\omega_{\mathrm{TO}}$), the group velocity of the LO phonon at the $\Gamma$ point, and the damping rate. From Equation~\eqref{eq:zerosofresponse}, we can immediately write the conductivity as
  \begin{equation}
  \sigma(\mathbf{q},\omega) = \frac{-2\iu\epsilon_0\epsilon_{\mathrm{env}}\omega}{q}\frac{\omega^2_{\mathbf{q},\mathrm{LO}}-\omega^2_{\mathrm{TO}}}{\omega^2_{\mathrm{TO}}-\omega^2-\iu\omega\tau^{-1}}.
  \end{equation}
 In this expression, $\omega_\mathrm{L}$, the LO phonon frequency prior to LO--TO splitting, has been re-named as $\omega_{\mathrm{TO}}$, the transverse optical phonon frequency, because in the absence of LO--TO splitting, they are degenerate. The RPA zeros $\omega_{\mathbf{q}}$ have also been renamed as $\omega_{\mathbf{q},\mathrm{LO}}$. This is done in order to make the form of the final results more closely resemble their 3D counterparts, in which the dielectric function is expressed in terms of the TO frequency (see for example Eq. (\ref{eq:3deps})).

 For small $q$,  the Born charges are (to lowest-order) constant, and so the LO phonon dispersion takes the form $\omega_{\mathbf{q},\mathrm{LO}}=\sqrt{\omega^2_{\mathrm{TO}}+2v_{\mathrm{g}}\omega_{\mathrm{TO}}q} \simeq \omega_{\mathrm{TO}} + v_{\mathrm{g}}q$, where $v_{\mathrm{g}}$, the LO phonon group velocity, is defined from microscopic parameters through the relation
 \begin{equation}
    v_{\mathrm{g}} = \frac{\Big|\hat{\mathbf{q}}\cdot \sum_\kappa \mathbf{Z}_\kappa\boldsymbol{\eta}_\kappa\Big|^2 }{4\epsilon_0 \epsilon_{\mathrm{env}}\omega_{\mathrm{TO}}\Omega}. 
\end{equation}
Thus, in the long wavelength limit, we have the following universal parameterization of the conductivity of a polar monolayer:
\begin{equation}
    \sigma(\omega) =  \frac{-4\iu\epsilon_0\epsilon_{\mathrm{env}}\omega\omega_{\mathrm{TO}}v_{\mathrm{g}}}{\omega^2_{\mathrm{TO}}-\omega^2-\iu\omega\tau^{-1}}.
    \label{eq:localsigma}
\end{equation}

We note that despite its appearance, $\sigma(\omega)$ does not depend on $\epsilon_{\mathrm{env}}$, as $v_{\mathrm{g}}$ has an opposite dependence on $\epsilon_{\mathrm{env}}$. From this relation, it follows that given the properties of the 2D phonons (from experiments or from \emph{ab initio} calculations), one can immediately specify the conductivity. Alternatively, from optical measurements (including far-field measurements) which allow one to extract the conductivity, it becomes possible to extract the group velocity of 2D LO phonons and thus the small-wavevector dispersion of those phonons.

Before moving on to analyze the electrodynamics of 2D phonon polaritons, we make three comments on lack of LO--TO splitting in 2D polar materials. The first is that this situation is in stark contrast to the situation of polar materials in 3D, which have a finite LO--TO splitting at the $\Gamma$ point. In the absence of such LO--TO splitting in 3D, there would be no frequency compatible with the existence of a phonon  polariton. On the contrary, we will show that in 2D, despite the absence of LO--TO splitting at the $\Gamma$ point, there persists a strongly confined evanescent electromagnetic mode with a high local density of states which in all respects is similar to a phonon polariton of a thin film, but is in fact the 2D LO phonon of the polar monolayer (thus the phrases ``phonon polariton'' and ``2D LO phonon'' may be used somewhat interchangeably as is the case in plasmonics where the terms ``plasmon polariton'' and ``2D plasmon'' are often used interchangeably).

The second comment is that much of what has been discussed here has a strong analogy with the theory of optical response in electron gases in 2D, and particularly the relation between plasmons in 2D and 3D. To elaborate on this analogy, we take Equation~\eqref{eq:zerosofresponse} in the case of a two-atom unit cell (such as hBN), and note that the term in the sum over Born charges can be written as $\Big|\hat{\mathbf{q}}\cdot \sum_\kappa \mathbf{Z}_\kappa\boldsymbol{\eta}_\kappa\Big|^2 \equiv \frac{Q_*^2}{M_*}$, with $Q_*$ being an effective charge and $M_*$ being an effective mass. Then, the LO--TO splitting can be written as $\omega^2_{\mathbf{q}} - \omega^2_{\mathbf{q},\mathrm{TO}} = \frac{Q_*^2}{2\epsilon_0\epsilon_{\mathrm{env}} M_*}q$. Now we note that the RHS is exactly the squared-frequency $\omega^2_{\mathbf{q}\mathrm{p}}$ for a plasma oscillation in a 2D gas of charged particles with charge $Q_*$ and mass $M_*$. To connect to LO--TO splitting in phonons, this squared frequency $\omega^2_{\mathbf{q}\mathrm{p}}$ can be thought of as the ``LP--TP'' splitting between longitudinal and transverse plasma oscillations. Of course, there are no transverse plasma oscillations due to the structure-less nature of the electron, and so ``$\omega_{\mathrm{TP}}$''  should be considered equal to zero. In the three-dimensional plasmon case, ``$\omega_{\mathrm{TP}} = 0$'', but the difference between the squared longitudinal and transverse plasma oscillation frequencies at zero-wavevector is non-zero and given by $\omega_{\mathrm{p}}^2$. In other words, the plasma frequency in electron gases is analogous to the LO--TO splitting in polar materials. The change in the dispersion of 3D versus 2D plasmons, like 3D versus 2D LO phonons, arises from the change in dimensionality of the Coulomb interaction from 3D to 2D.  This analogy between phononic and plasmonic behavior as a function of dimension is illustrated schematically in Figures~\ref{fig:1}(b,c) in order to help unify the understanding of 2D plasmonics and phononics. These considerations should also extend to the one-dimensional case.

\addition{%
The third comment is that the considerations of this section can be extended to few-layer systems, such as hBN bilayers, which may still be considered 2D for polariton wavelengths large compared to the bilayer thickness. The considerations may also be extended to 1D polar materials, such as BN nanotubes. 
For bilayers, as was discussed in Ref.~\citenum{sohier2017breakdown}, one may describe the LO--TO splitting in terms of Eq.~\eqref{eq:zerosofresponse}, with the 2D Coulomb interaction, but with eigendisplacements appropriate to the few layer system.
For a small number of layers $N$, \FIXME{Mention scaling of $|\hat{\mathbf{q}}\cdot\ldots|^2 = NQ_*^2/M_*$} this should enhance the LO--TO splitting by a factor of $N$. 
To go beyond an effectively 2D description of a monolayer, and make the transition from monolayer to bulk, we may make use of the quasi-2.5D Coulomb interaction~\cite{thygesen2017calculating} in Eq.~\eqref{eq:zerosofresponse}. 
In that case, for a film of effective thickness $t$, the interaction is approximately $V_{\text{Q2.5D}}(q) = \frac{e^2}{\epsilon_0 tq^2}\big(1-\frac{2}{qt}e^{-\frac{qt}{2}}\sinh\frac{qt}{2} \big)$. 
This form interpolates between the 2D and 3D interactions in the limits $qt \rightarrow 0$ and $qt \rightarrow \infty$. 
For sufficiently large thickness and finite-$q$, the 2.5D Coulomb interaction scales as $t^{-1}q^{-2}$. 
Noting the $q^2$ factor in Eq.~\eqref{eq:zerosofresponse}, as well as the $N$-dependence of the quantity involving eigendisplacements, one finds an LO--TO splitting independent of thickness, as expected for bulk samples.
}

\addition{%
For 1D systems, beyond using the eigendisplacements appropriate to the 1D system, one must also use the 1D Coulomb interaction, cutoff by a length $R$ which for a nanotube, would be its radius. 
In that case, the 1D Coulomb interaction is $e^2K_0(qr)/2\pi\epsilon_0$, where $K_0$ is the (zeroth order) modified Bessel function of the second kind. 
Thus, the $q$-dependence of the LO--TO splitting is $q^2K_0(qR)$. This is in close analogy with the relation for the plasma frequency of 1D plasmons, as expected from the considerations of the previous paragraph. 
As regards the optical properties, the optical conductivity is still governed by Eq.~\eqref{eq:2dsusceptibility}.
}

\section{Electrodynamics of optical phonons in two-dimensions}
To relate the conductivity function  to the electromagnetic modes supported by a polar monolayer, we solve Maxwell's equations for an evanescent electromagnetic mode supported by a surface with conductivity $\sigma$. We consider the monolayer to be sandwiched by a superstrate of permittivity $\epsilon_+$ and a substrate of permittivity $\epsilon_-$. To strip the analysis to its bare essentials, we consider optical phonon response with in-plane isotropy in the long-wavelength limit arising from in-plane LO oscillations. A relevant example of a system where these conditions are satisfied is in a hexagonal boron nitride monolayer (see Figure~\ref{fig:1}(a) for schematic atomic structure). In a monolayer geometry with translation invariance and in-plane isotropy, the solutions of Maxwell's equations can be decomposed into transverse magnetic (TM) and transverse electric (TE) parts, where the magnetic or electric field respectively is transverse to the in-plane wavevector of the mode. In practice, it is the TM mode which is associated with highly confined electromagnetic waves.   We consider without loss of generality a TM mode with wavenumber $q$ along the $x$-direction in the monolayer and magnetic field $H(z)\e^{\iu qx-\iu\omega t}$ along the $y$-direction of the monolayer. The direction transverse to the monolayer is denoted as $z$.  With these definitions in place, the Maxwell equation satisfied by the magnetic field is 
\begin{equation}\label{eq:2dmaxwell}
    \bigg(-\frac{\dd^2}{\dd{}z^2}+q^2-\epsilon_{\pm}\frac{\omega^2}{c^2} \bigg)H(z) = 0,
\end{equation}
where $\epsilon_+$ applies for $z>0$ and $\epsilon_-$ applies for $z<0$.
We consider a solution of the form $H(z) = h_\pm\e^{\mp\kappa_\pm z}$ with $\kappa_\pm = \sqrt{q^2-\epsilon_{\pm} \frac{\omega^2}{c^2}}$ with $\pm$ corresponding to $\pm z > 0$ respectively.  The boundary condition on the magnetic field is $h_+ - h_- = -K_x = -\sigma E_x$ where $\mathbf{K}$ is the surface current density, and $\mathbf{E} = -\frac{1}{i\omega\epsilon}\nabla\times\mathbf{H}$ is the electric field. This condition enforces $h_+-h_- = \frac{\sigma}{i\omega\epsilon_+}\kappa_+h_+$. 
Continuity of the electric field in the $x$ direction enforces ${\epsilon_-}/{\epsilon_+} = -{\kappa_-h_-}/{\kappa_+h_+}$. 
Combining the two conditions, we obtain the usual dispersion equation for the TM mode of a polarizable 2D monolayer, namely $\epsilon_+/\kappa_+ + \epsilon_-/\kappa_- = \frac{\sigma}{i\omega}$. Given the deeply subwavelength nature of 2D phonon polaritons, \ie since $q\gg\omega/c$ such that $\kappa_{\pm}\simeq q$, the dispersion equation can be reduced to its quasistatic limit without consequential loss of accuracy:
\begin{equation}
 q=\frac{2i\omega\epsilon_0\epsilon_{\mathrm{env}}}{\sigma},
    \label{eq:dispeq}
\end{equation}
with $\epsilon_{\mathrm{env}}\equiv (\epsilon_++\epsilon_-)/2$. This condition, as can be seen from the relation between $\Pi$ and $\sigma$, is equivalent to $\epsilon_{\mathrm{RPA}}=0$. Thus, the condition for phonon polaritons coincides precisely with the condition for longitudinal optical phonons.  One of the main results of our manuscript is that despite the lack of LO--TO splitting at the $\Gamma$ point, there nevertheless exists a strongly confined evanescent mode in many respects similar to the phonon polaritons of thin films. We now analyze the dispersion relation of phonon polaritons in a specific material, hexagonal boron nitride, in more detail, showing the possibility of highly confined electromagnetic modes with a large local density of states.

\begin{figure}[t]
    \includegraphics[width=0.95\linewidth]{fig1.pdf}
    \caption{%
        \textbf{LO phonons as the basic electromagnetic waves of a polar monolayer.} (a) Schematic structure of a polar monolayer such as hexagonal boron nitride. (b) Properties of LO and TO phonons in 3D and 2D. In 3D, there is a finite LO--TO splitting at zero wavevector, while in 2D there is none. Despite this, the 2D LO phonon plays the role of the phonon polariton in 3D and thin films. (c) Analogous physics appears in electron gases in 3D and 2D, where the 3D plasma frequency is similar to the 3D LO--TO splitting. In 2D, the plasma frequency at zero wavevector is zero, but the electromagnetic physics is determined by the dispersion of 2D plasmons, which replace the plasmon polariton of bulk and thin films.
        \label{fig:1}
        }
\end{figure}~

\begin{figure*}[t]
    \includegraphics[width=.7\linewidth]{figure2hBN.pdf}
    \caption{%
        \textbf{Properties of phonon polaritons in a monolayer of hBN compared to bulk.} (a)~Real and imaginary parts of the conductivity of 2D hBN for different values of the loss-rate. (b)~Propagation quality factor, which measures the number of wavelengths of propagation of the 2D phonon polariton. (c)~Dispersion relation of phonon polaritons in the monolayer (black) and thin films whose thicknesses are taken to be 1, 2, 4, and 8 interlayer spacings in hBN. For the thin-film, the fundamental mode (dashed) and the first higher-order mode (dash-dotted) are plotted. 
        \label{fig:2}
        }
\end{figure*}

In Figure ~\ref{fig:2}(a), we present the conductivity of 2D hBN of Equation~\ref{eq:localsigma}, using parameters from Ref.~\citenum{sohier2017breakdown} calculated from density functional theory within the local density approximation. In this plot $\epsilon_{\mathrm{env}}=1,$ $\omega_{\mathrm{TO}} = $ \SI{1387}{\per\cm} and $v_{\mathrm{g}} = 1.2\times 10^{-4}c$, with $c$ the speed of light in vacuum.

From the conductivity, the dispersion relation of phonon polaritons on an infinite sheet is given by $q={2\iu\omega\epsilon_0\epsilon_{\mathrm{env}}}/{\sigma(\omega)}$. The dispersion, assuming $\tau = \infty$ and $\epsilon_{\mathrm{env}}=1$, is shown in Figure~\ref{fig:2}(c) (black line). A key figure of merit for applications involving the propagation of phonon polaritons, is the propagation quality factor, defined by ${\Re q}/{\Im q} = {\Im\sigma(\omega)}/{\Re\sigma(\omega)}$, which is shown in Figure~\ref{fig:2}(b).

For monolayer hBN, the wavevector grows very rapidly with frequency, due to the extremely low group velocity of 2D LO phonons, which is a remarkable four orders of magnitude slower than the speed of light. In particular, at frequencies of \SI{1450}{\per\cm}, the phonon polariton has a wavelength of about \SI{15}{\nm}, significantly shorter than any phonon polariton measured so far, and, similarly shorter than any plasmonic wavelength, even in graphene. In fact, this short a wavelength well-below that of any polariton in current scattering near-field microscopy (SNOM) measurements. The 2D phonon polariton could in principle be measured by SNOM closer to the TO frequency, where confinement is smaller; unfortunately, as shown in Figure~\ref{fig:2}(b), near the TO frequency, dissipation is far higher (and corresponding propagation quality factors $\Re q/\Im q$ far lower) due to large $\Re\sigma$ (or, equivalently, large $\Im\epsilon_{\mathrm{RPA}}$).

>>>>

\documentclass[aps,prb,twocolumn, final,
	           groupedaddress,superscriptaddress,
               amsfonts,amssymb,amsmath,floatfix,
	           citeautoscript]{revtex4-1}

@@

% ----- COMMENTS AND META-ANNOTATIONS -----
\usepackage{textcomp} % for \textrightarrow
\usepackage{xifthen}
\usepackage{etoolbox}
\newboolean{togglecomments}
\newboolean{togglechanges}

% toggle to true to see comments (otherwise hidden)
\setboolean{togglecomments}{true}  
% toggle to false to see mixed versions (otherwise edits are shown exclusively)
\setboolean{togglechanges}{false}

\newcommand{\comment}[2]{%
    \ifbool{togglecomments}%
    {\textcolor{blue!70!black}{\small\textsf{%
    \textsuperscript{\textsc{\textsf{\MakeLowercase{#1}}}}%
    [#2]}}} % if true, show comments
    {}}     % if false, do nothing
\newcommand{\swap}[2]{\ifbool{togglechanges}
    {#2}  % TC-only version
    {\textcolor{red!70!black}{[#1]}\text{\textrightarrow{}}\textcolor{green!50!black}{[#2]}}}
\newcommand{\remove}[1]{\ifbool{togglechanges}
    {}    % TC-only version
    {\textcolor{red!70!black}{#1}}}
\newcommand{\inset}[1]{\ifbool{togglechanges}
    {{}#1}  % TC-only version
    {\textcolor{green!50!black}{{}#1}}}
\makeatletter
\newcommand{\citeremind}[1]{%
	\unskip%
    \textcolor{blue!75!black!80!yellow}{${}^\blacksquare$%
	\ifthenelse{\isempty{#1}}{}{\textsuperscript{\tiny\textsf{#1}}}%
	}\xspace}
\makeatother
\newcommand{\FIXME}[1]{\textcolor{red!70!black}{[#1]}}
\newcommand{\optional}[1]{\textcolor{yellow!70!gray}{#1}}

@@

% --- TITLE & AUTHORSHIP ---
\title{Supporting Information for: Phonon polaritonics in two-dimensional materials}

@@

% --- ABSTRACT ---

@@

\begin{figure*}[h]
    \includegraphics[width=0.6\linewidth]{enhancement.png}
    \caption{%
        \textbf{Spontaneous emission enhancement in hBN thin films and monolayer.} Spontaneous emission enhancement for a dipole emitter 5 nm away from the surface of thin films of bulk hBN as a function of emission frequency for film thicknesses of 100 nm, 10 nm, 1 nm, and 0.33 nm, as well as monolayer hBN with dispersion relation and conductivity given in the main text. The hBN in all cases is surrounded by air. The TO frequency  taken for the bulk is 1387 cm$^{-1}$.
        \label{fig:s1}
        }
\end{figure*}~
